preciseTAD: A machine learning framework for precise TAD boundary prediction at base pair resolution

### 2020-07-11 Manuscript/manuscript.Rmd
addressed some ???

[??? There is a big gap there. Describe the construction of dependent vector Y and predictor matrix _for each base_;!!!talk about in meeting]

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* [??? We need to define what to italicize. Check below, make consistent throughout the manuscript;!!!talk about in meeting] (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* [??? Again, I expected italics, but decide and correct;!!!talk about in meeting] and *plotHeatmap* tools offered by *deepTools* version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and Peakachu loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/red for Arrowhead/Peakachu, green for preciseTAD?;!!!can't be done for peakachu because we only have 1 resolution (10kb)]

@@

![**preciseTAD boundaries are more conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models.](Figures/figure6_venn_diagrams.png) 
[??? What is the difference between leftmost Venn diagrams on panel A?] [??? What are the numbers?;!!!talk in meeting]

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7 [??? Merge them into one Excel with two sheets - same message]). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary [??? Is it because of class imbalance?]. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well [??? So, another class imbalance problem? Define class imbalance message early, why you investigated 5 other metrics]. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it?]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [??? The logic is messed when reading this paragraph.]

>>>>

[??? There is a big gap there. Describe the construction of dependent vector Y and predictor matrix _for each base_;!!!talk about in meeting; specify that each base pair has own prob; don't use "base pair" only "base" or "base level"]

First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, DBSCAN [??? We need to define what to italicize. Check below, make consistent throughout the manuscript;!!!talk about in meeting] (version 1.1-5) was applied to the matrix of pairwise genomic distances between base pairs with $p_{n_{i}}=1$. This was to avoid the inflation of predicted boundaries that can occur naturally when dealing with such fine resolution, namely, due to spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts*. The *eps* parameter was assigned as the respective data resolution. The heuristic being that density-reachable base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding *MinPts* parameter was set to recommended value of 3, representing $1+dim(data)$ [@hahsler2019dbscan]. The resulting clusters of highly predictive base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR). To precisely identify a single base pair among each PTBR, preciseTAD then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. The corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative and biologically meaningful base pair coordinate within each clustered PTBR.

@@

Next, we assessed the biological significance of our predicted boundaries with respect to the enrichment profiles of several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the computeMatrix [??? Again, I expected italics, but decide and correct;!!!talk about in meeting] and plotHeatmap tools offered by deepTools version 2.0 [@ramirez2016deeptools2] via integration with Galaxy version 3.3.2.0.0 [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum test. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using Jaccard index defined as

@@

![**preciseTAD boundaries are enriched in CTCF sites.** (A) The location of preciseTAD (green) vs. ARROWHEAD (blue) boundaries on GM12878 (chr 1: 70600000-71530000). The black line represents the probability of each base being a TAD boundary. The green vertical lines represent the *preciseTAD boundary points* (PTBP). A zoomed-in portion of the genome shows a yellow-highlighted cluster of bases with high probability of being a boundary, the *preciseTAD boundary region (PTBR). (B) Heatmaps comparing the enrichment of CTCF signal around preciseTAD and ARROWHEAD TAD boundaries, and (C) around preciseTAD and Peakachu loop boundaries. (D) $Log_{2}$ genomic distances from boundaries to nearest CTCF sites. (E) Pairwise Jaccard indices shows better overlap among *preciseTAD*-detected boundaries across resolutions (upper triangular) as compared to ARROWHEAD (lower triangular).](Figures/figure5_enrichment_and_log2dist_ctcf.png) [??? Panel E - do we have it for Peakachu? - stretched, can be put in figure. Can you color triangulars blue/green for Arrowhead/preciseTAD, green for preciseTAD?;!!!can't be done for peakachu because we only have 1 resolution (10kb)]

@@

![**preciseTAD boundaries are more conserved between the GM12878 and K562 cell lines.** (A) Venn diagrams illustrating the overlap between GM12878- and K562-specific predicted and called boundaries. (B) Venn diagrams illustrating the agreement between boundaries predicted using ARROWHEAD- and Peakachu-trained models.](Figures/figure6_venn_diagrams.png) 
[??? What is the difference between leftmost Venn diagrams on panel A?] [??? What are the numbers?;!!!talk in meeting; add indication of what ground truth is for preciseTAD]

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Supplementary Tables 6-7 [??? Merge them into one Excel with two sheets - same message]). The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary [??? Is it because of class imbalance?; add detail]. AUROC captures how a model generally performs accross different thresholds. However, it doesn't place more emphasis on one class over the other, so it does not reflect the minority class well [??? So, another class imbalance problem? Define class imbalance message early, why you investigated 5 other metrics]. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be very sensitive towards disparate values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance and also omits from its calculation TN values [??? So why didn't you use it? for MD]. All of these are important considerations to make when choosing a performance metric. For these reasons we opted to report the balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data re-sampling. [??? The logic is messed when reading this paragraph.]
